---
title: "BACI Trade Network Analysis — Full Pipeline"
subtitle: "Network-Level Indicators, Node-Level Causality, and Structural Visualization"
author: "Dénes Kiss"
output:
  html_document:
    highlight: haddock
    theme: lumen
    toc: yes
    number_sections: true
    toc_float:
      collapsed: no
  html_notebook:
    fig_caption: yes
    highlight: haddock
    theme: lumen
    toc: yes
    number_sections: true
    toc_float:
      collapsed: no
  pdf_document:
    keep_tex: yes
    latex_engine: lualatex
    toc: yes
    number_sections: true
  word_document:
    toc: yes
---

# Overview

This script reproduces the full BACI-CEPII trade network analysis pipeline used in
the dissertation *"Reservoir Nodes - Identifying Drivers and Structural Change in
Trade and Mobility Networks."* It covers three stages:

1. **Network-level indicators** — 27 structural metrics per product per year,
   computed for both value-weighted (BACI) and volume-weighted (BACI volume) data.
2. **Node-level indicators and causality** — country-level centrality measures
   per product group, followed by pairwise Granger causality testing to identify
   causal communities of products.
3. **Visualization** — temporal clustering of network indicators and a world map
   of node-level scores.

Set `PATH` below to the local folder containing the BACI zip file before running.

# Load necessary libraries

```{r, warning=FALSE, message=FALSE}
library(igraph)
library(nda)
library(readr)
library(dplyr)
library(brainGraph)
library(robin)
library(lmtest)
library(countrycode)
library(maps)
library(ggplot2)
```

# Specify necessary functions

```{r, warning=FALSE, message=FALSE}
# Percolation function — simulates sequential vertex removal to trace
# the size of the giant component (used for resilience indicators)
percolate <- function(g, size, d) {
  giant <- vector()
  c <- components(g)
  giant[1] <- max(c$csize)

  names(d) <- 1:length(d)
  d <- sort(d, decreasing = TRUE)
  vital <- as.integer(names(d[1:size]))

  for (i in 1:size) {
    c <- components(delete_vertices(g, vital[1:i]))
    giant[i + 1] <- max(c$csize)
  }
  return(giant)
}

# Replaces NA/NULL values with 0 and pads to length N
fillzeros <- function(v, N = length(v)) {
  v <- as.vector(v)
  V <- as.vector(matrix(0, nrow = 1, ncol = N))
  for (i in 1:N) {
    if (is.na(v[i]) || is.null(v[i])) {
      # leave as 0
    } else {
      V[i] <- v[i]
    }
  }
  return(V)
}
```

# Set data path

```{r, warning=FALSE, message=FALSE}
# Set your local path to the folder containing the BACI zip file
if (.Platform$OS.type == "unix") {
  PATH <- "~/Library/Group Containers/G69SCX94XU.duck/Library/Application Support/duck/Volumes/S/DATABASES/CEPII/"
} else {
  PATH <- "S:/DATABASES/CEPII/"
}

FILE <- paste(PATH, "BACI_HS92_V202201.zip", sep = "")
LIST <- unzip(zipfile = FILE, list = TRUE)
LIST <- LIST[grep("BACI", LIST$Name), 1]
```

---

# Section 1 — Network-Level Indicators

This section computes 27 network-level indicators per product per year, for both
value-weighted and volume-weighted trade flows. Indicators correspond directly to
Tables 3.1 and 3.2 in the dissertation.

## 1.1 Value-weighted indicators (BACI)

```{r, warning=FALSE, message=FALSE}
NETWORK_COLS <- c(
  "Product", "Year", "E_Connectivity", "V_Connectivity", "Assortativity",
  "Mean_coreness_all", "Mean_coreness_in", "Mean_coreness_out",
  "Clustering_global", "Clustering_local_avg", "Clustering_weighted",
  "C_betweenness", "C_evcent", "RC_phi", "RC_Ek", "RC_Nk",
  "Robustness_random", "Gamma", "Weak_conn_comp", "Strong_conn_comp",
  "Biconn_comp", "Cohesion", "Res_rand", "Res_deg", "Reciprocity",
  "E_asymm", "N_asymm"
)

compute_network_indicators <- function(DATA, weight_col) {
  U <- unique(DATA$k)
  N <- length(U)
  OUTDATA <- as.data.frame(matrix(0, nrow = N, ncol = 27))
  colnames(OUTDATA) <- NETWORK_COLS
  rownames(OUTDATA) <- U

  for (i in 1:N) {
    g <- graph.edgelist(as.matrix(DATA[DATA$k == U[i], c(2, 3)]))
    E(g)$weight <- as.matrix(DATA[DATA$k == U[i], weight_col])

    OUTDATA[U[i], 1] <- U[i]
    OUTDATA[U[i], 2] <- DATA[1, 1]
    OUTDATA[U[i], 3] <- edge_connectivity(g)
    OUTDATA[U[i], 4] <- vertex_connectivity(g)
    OUTDATA[U[i], 5] <- assortativity.degree(g)
    OUTDATA[U[i], 6] <- mean(coreness(g, mode = "all"), na.rm = TRUE)
    OUTDATA[U[i], 7] <- mean(coreness(g, mode = "in"), na.rm = TRUE)
    OUTDATA[U[i], 8] <- mean(coreness(g, mode = "out"), na.rm = TRUE)
    OUTDATA[U[i], 9] <- transitivity(g, type = "global")
    OUTDATA[U[i], 10] <- transitivity(g, type = "localaverage")
    OUTDATA[U[i], 11] <- mean(transitivity(g, type = "weighted"), na.rm = TRUE)
    OUTDATA[U[i], 12] <- centralization.betweenness(g)$centralization
    OUTDATA[U[i], 13] <- centralization.evcent(g)$centralization
    OUTDATA[U[i], 14] <- rich_club_coeff(g, weighted = TRUE)$phi
    OUTDATA[U[i], 15] <- rich_club_coeff(g, weighted = TRUE)$Ek
    OUTDATA[U[i], 16] <- rich_club_coeff(g, weighted = TRUE)$Nk
    OUTDATA[U[i], 17] <- mean(robustness(g, type = "vertex", measure = "degree", N = 1000)$comp.pct, na.rm = TRUE)
    OUTDATA[U[i], 18] <- ifelse(fit_power_law(degree(g))$KS.p > 0.05, fit_power_law(degree(g))$alpha, 0)
    OUTDATA[U[i], 19] <- components(g, mode = "weak")$no
    OUTDATA[U[i], 20] <- components(g, mode = "strong")$no
    OUTDATA[U[i], 21] <- biconnected.components(g)$no
    OUTDATA[U[i], 22] <- cohesion(g)
    OUTDATA[U[i], 23] <- mean(percolate(g, vcount(g) / 2, d = sample(V(g), vcount(g) / 2)), na.rm = TRUE)
    OUTDATA[U[i], 24] <- mean(percolate(g, vcount(g) / 2, d = degree(g)), na.rm = TRUE)
    OUTDATA[U[i], 25] <- reciprocity(g)
    OUTDATA[U[i], 26] <- mean(abs(as.matrix(get.adjacency(g)) - t(as.matrix(get.adjacency(g)))) /
      (as.matrix(get.adjacency(g)) + t(as.matrix(get.adjacency(g)))), na.rm = TRUE)
    OUTDATA[U[i], 27] <- sum(abs(rowSums(as.matrix(get.adjacency(g)), na.rm = TRUE) -
      colSums(as.matrix(get.adjacency(g)), na.rm = TRUE))) /
      sum(abs(rowSums(as.matrix(get.adjacency(g)), na.rm = TRUE) +
        colSums(as.matrix(get.adjacency(g)), na.rm = TRUE)))
  }
  return(OUTDATA)
}

nL <- length(LIST)
for (j in 1:nL) {
  unzip(zipfile = FILE, files = LIST[j])
  DATA <- read_csv(LIST[j])

  # Value-weighted (column 6 = value)
  OUTDATA_VAL <- compute_network_indicators(DATA, weight_col = 6)
  write.table(OUTDATA_VAL, file = paste("OUT_NET_", DATA[1, 1], ".csv", sep = ""),
              sep = ";", row.names = FALSE)

  # Volume-weighted (column 5 = quantity)
  OUTDATA_VOL <- compute_network_indicators(DATA, weight_col = 5)
  write.table(OUTDATA_VOL, file = paste("OUT_NETVOL_", DATA[1, 1], ".csv", sep = ""),
              sep = ";", row.names = FALSE)

  unlink(LIST[j])
}
```

## 1.2 Assemble 3D arrays (product × indicator × year)

```{r, warning=FALSE, message=FALSE}
assemble_3d_array <- function(prefix) {
  DIR <- dir()
  DIR <- DIR[grep(prefix, DIR)]
  Tables <- list()
  products <- NULL

  for (i in 1:length(DIR)) {
    tmp <- read_delim(DIR[[i]], delim = ";", escape_double = FALSE, trim_ws = TRUE)
    if (i > 1) {
      products <- full_join(products, tmp[1])
    } else {
      products <- tmp[1]
    }
    Tables[[i]] <- read.csv(DIR[[i]], sep = ";", row.names = 1, header = TRUE)
  }

  ar <- array(NA,
    dim = c(nrow(products), ncol(Tables[[1]]), length(1995:2020)),
    dimnames = list(as.matrix(products), as.matrix(colnames(Tables[[1]])), c(1995:2020))
  )
  for (i in 1:length(DIR)) {
    ar[dimnames(ar)[[1]], , i] <- as.matrix(Tables[[i]][dimnames(ar)[[1]], ])
  }
  return(ar)
}

ar_value <- assemble_3d_array("OUT_NET_")
ar_volume <- assemble_3d_array("OUT_NETVOL_")

save(ar_value, ar_volume, file = "BACINET3D.RData")
```

## 1.3 Temporal clustering of network indicators (GNDA)

```{r, warning=FALSE, message=FALSE}
# Spearman partial-correlation-based NDA graph — value-weighted, asymmetry indicator (col 26)
NDA_value <- ndr(na.omit(t(na.omit(ar_value[, 26, ]))), cor_method = 2, cor_type = 2)
plot(NDA_value, cuts = 0.45)

# Same for volume-weighted
NDA_volume <- ndr(na.omit(t(na.omit(ar_volume[, 26, ]))), cor_method = 2, cor_type = 2)
plot(NDA_volume, cuts = 0.45)
```

---

# Section 2 — Node-Level Indicators and Causality

This section computes country-level centrality indicators for each two-digit
product group (HS2) and tests pairwise Granger causality between countries'
structural trajectories, identifying causal communities (Table 3.1, Section 4.2
of the dissertation).

## 2.1 Compute node-level indicators per product group

```{r, warning=FALSE, message=FALSE}
NODE_INDICATOR_NAMES <- c("EV", "PR", "CN", "CL", "BN", "DA", "DI", "DO", "NA")

nL <- length(LIST)
for (j in 1:nL) {
  unzip(zipfile = FILE, files = LIST[j])
  DATA <- read_csv(LIST[j])
  U <- unique(substr(DATA$k, 1, 2))
  N <- length(U)
  M <- max(max(DATA[, 2]), max(DATA[, 3]))

  outdata_list <- list()
  for (name in NODE_INDICATOR_NAMES) {
    df <- as.data.frame(matrix(0, nrow = N, ncol = M + 2))
    rownames(df) <- U
    colnames(df) <- c("Product", "Year", 1:M)
    outdata_list[[name]] <- df
  }

  for (i in 1:N) {
    sub <- DATA[substr(DATA$k, 1, 2) == U[i], ]
    g <- graph.edgelist(as.matrix(sub[, c(2, 3)]))
    E(g)$weight <- abs(as.matrix(fillzeros(as.matrix(sub[, 5]))))

    vals <- list(
      EV = eigen_centrality(g, weights = E(g))$vector,
      PR = page_rank(g, weights = E(g))$vector,
      CN = coreness(g),
      CL = harmonic_centrality(g, weights = E(g)),
      BN = betweenness(g, weights = E(g)),
      DA = degree(g, mode = "all"),
      DI = degree(g, mode = "in"),
      DO = degree(g, mode = "out"),
      NA_ = strength(g, weights = E(g), mode = "out") -
        strength(g, weights = E(g), mode = "in") / strength(g, weights = E(g), mode = "all")
    )

    for (name in NODE_INDICATOR_NAMES) {
      key <- ifelse(name == "NA", "NA_", name)
      outdata_list[[name]][U[i], 1] <- U[i]
      outdata_list[[name]][U[i], 2] <- DATA[1, 1]
      outdata_list[[name]][U[i], -(1:2)] <- fillzeros(vals[[key]], M)
    }
  }

  for (name in NODE_INDICATOR_NAMES) {
    write.table(outdata_list[[name]],
      file = paste("OUT_NODE_", name, "_", DATA[1, 1], ".csv", sep = ""),
      sep = ";", row.names = FALSE
    )
  }
  unlink(LIST[j])
}
```

## 2.2 Load country codes and assemble node-level 4D arrays

```{r, warning=FALSE, message=FALSE}
CC <- unzip(zipfile = FILE, list = TRUE)
CC <- CC[grep("country", CC$Name), 1]
unzip(zipfile = FILE, files = CC)
CCodes <- read_csv(CC)
unlink(CC)

assemble_node_array <- function(indicator) {
  DIR <- dir()
  DIR <- DIR[grep(paste0("OUT_NODE_", indicator, "_"), DIR)]
  Tables <- list()
  products <- NULL

  for (i in 1:length(DIR)) {
    tmp <- read_delim(DIR[[i]], delim = ";", escape_double = FALSE, trim_ws = TRUE)
    if (i > 1) {
      products <- full_join(products, tmp[1])
    } else {
      products <- tmp[1]
    }
    Tables[[i]] <- read.csv(DIR[[i]], sep = ";", row.names = 1, header = TRUE)
  }

  ar <- array(NA,
    dim = c(nrow(products), ncol(Tables[[1]]), length(1995:2020)),
    dimnames = list(as.matrix(products), as.matrix(colnames(Tables[[1]])), c(1995:2020))
  )
  for (i in 1:length(DIR)) {
    ar[dimnames(ar)[[1]], , i] <- as.matrix(Tables[[i]][dimnames(ar)[[1]], ])
  }
  ar <- ar[, paste("X", CCodes$country_code, sep = ""), ]
  dimnames(ar)[[2]] <- CCodes$iso_3digit_alpha
  return(ar)
}

ar_EV <- assemble_node_array("EV")
ar_PR <- assemble_node_array("PR")
ar_CN <- assemble_node_array("CN")
ar_CL <- assemble_node_array("CL")
ar_BN <- assemble_node_array("BN")
ar_DA <- assemble_node_array("DA")
ar_DI <- assemble_node_array("DI")
ar_DO <- assemble_node_array("DO")
ar_NA <- assemble_node_array("NA")

save(ar_EV, ar_PR, ar_CN, ar_CL, ar_BN, ar_DA, ar_DI, ar_DO, ar_NA,
     file = "BACINODE4D.RData")
```

## 2.3 Temporal classification of node trajectories (GNDA)

```{r, warning=FALSE, message=FALSE}
# Example: Product group 81 (base metals, tungsten) — index in array depends on
# product ordering; adjust Coef to the row index of the product group of interest.
Coef <- 79 # Product group 81

data_da <- as.data.frame(na.omit(t(na.omit(ar_DA[Coef, , ]))))
tdata_da <- na.omit(as.data.frame(t(data_da)))

NDA_node <- ndr(data_da, cor_method = 2) # Spearman's correlation distance
tdata_da$NDA <- NDA_node$membership

for (i in 1:NDA_node$factors) {
  matplot(t(tdata_da[tdata_da$NDA == i, -ncol(tdata_da)]),
    type = "l", xlab = "years", ylab = "Degree centrality",
    main = "Time-series data", xaxt = "n"
  )
  nn <- nrow(as.matrix(rownames(tdata_da[tdata_da$NDA == i, ])))
  legend(0, max(tdata_da[tdata_da$NDA == i, -ncol(tdata_da)]),
    legend = rownames(tdata_da[tdata_da$NDA == i, ]),
    col = seq_len(nn), cex = 0.8, fill = seq_len(nn), lty = seq_len(nn)
  )
  axis(1, at = 1:length(1995:2020), labels = 1995:2020, cex.axis = 1.0)
}

# Cluster centroid plot
center <- matrix(0, length(1995:2020), NDA_node$factors)
colnames(center) <- paste("NDA_CLUST", 1:NDA_node$factors)
for (i in 1:NDA_node$factors) {
  center[, i] <- colMeans(tdata_da[tdata_da$NDA == i, -ncol(tdata_da)])
}
matplot(center, type = "l", xlab = "years", ylab = "Degree centrality",
        main = "Time-series data", xaxt = "n", lwd = 3.0)
legend(0, max(center), legend = colnames(center),
       col = seq_len(NDA_node$factors), cex = 0.8,
       fill = seq_len(NDA_node$factors), lty = seq_len(NDA_node$factors))
axis(1, at = 1:length(1995:2020), labels = 1995:2020, cex.axis = 1.0)
```

## 2.4 Granger causality between country trajectories

```{r, warning=FALSE, message=FALSE}
compute_causality_matrix <- function(indicator_array, product_index) {
  data <- t(indicator_array[product_index, , ])
  N <- ncol(data)
  CGM <- matrix(0, ncol = N, nrow = N)
  rownames(CGM) <- colnames(data)
  colnames(CGM) <- colnames(data)

  for (i in 1:(N - 1)) {
    for (j in (i + 1):N) {
      p_ij <- try(grangertest(data[, i], data[, j], 1)[2, 4], silent = TRUE)
      p_ji <- try(grangertest(data[, j], data[, i], 1)[2, 4], silent = TRUE)
      CGM[i, j] <- ifelse(as.numeric(p_ij) < 0.001, 1, 0)
      CGM[j, i] <- ifelse(as.numeric(p_ji) < 0.001, 1, 0)
    }
  }
  return(CGM)
}

# Example: causality graph for product group at index Coef, using betweenness trajectories
causality_graph <- compute_causality_matrix(ar_BN, Coef)
```

---

# Section 3 — Visualization

## 3.1 Map node-level scores to a world map

```{r, warning=FALSE, message=FALSE}
plot_node_indicator_map <- function(indicator_array, product_index, year_label) {
  # Extract values for one product group and one year, indexed by ISO3 country code
  values <- indicator_array[product_index, , year_label]
  map_data_df <- data.frame(
    iso_a3 = names(values),
    value = as.numeric(values)
  )
  map_data_df$country <- countrycode(
    sourcevar = map_data_df$iso_a3,
    origin = "iso3c", destination = "country.name", warn = FALSE
  )
  map_data_df <- na.omit(map_data_df)

  world_map <- map_data("world")
  merged <- merge(world_map, map_data_df, by.x = "region", by.y = "country", all.x = TRUE)

  ggplot(data = merged, aes(x = long, y = lat, group = group, fill = value)) +
    geom_polygon(color = "white", linewidth = 0.2) +
    scale_fill_gradient(low = "white", high = "red", na.value = "grey90") +
    coord_map("robinson") +
    theme_void() +
    labs(
      title = paste("Node-Level Indicator — Product Group", Coef, "—", year_label),
      fill = "Value"
    )
}

plot_node_indicator_map(ar_EV, Coef, "2020")
```

---

# Session info

```{r}
sessionInfo()
```
